TrenchR: an R package for transparent environmental and ecological biophysics

### 2022-06-28 Manuscript_full.Rmd
render
first set of edits
minor typos, bringing lines of the markdown doc together, rendering all three doc types

1.  Much understanding of organimsal responses to climate change and
    variability relies on the assumption that body temperatures are
    equal to temporally averaged air temperatures high above the ground.
    However, most organisms live near the ground and acute exposure to
    solar and thermal radiation and thermal extremes can substantially
    elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating
    Environmental Change into organismal responses. The package includes
    microclimate models to vertically scale weather station data to
    organismal heights. The biophysical modelling tools include both
    general models for heat flows and specific models to predict body
    temperatures for a variety of ectothermic taxa. Additional functions
    model and temporally partition air and soil temperatures and solar
    radiation. We offer utility functions to aid in estimating the
    organismal and environmental parameters needed for biophysical
    ecology.
3.  TrenchR focuses on simple and modular functions so users can create
    transparent and flexible models for biophysical applications.
4.  The package aims to introduce and enable microclimate and
    biophysical modelling to improve ecological and evolutionary
    forecasting. We further this aim through a series of educational
    modules that introduce the field of biophysical ecology.

@@

**Figure 1** A) Microclimate models scale air and soil temperatures and windspeeds from sensor to organismal heights. B) Biophysical models balance heat exchanges between organisms and their environment to estimate body temperatures [@porter1973behavioral]. A thermal image depicts how grasshopper body temperatures vary substantially from vegetation, air, and ground temperatures due to heat exchange.

@@

## Methods and features

The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to
promote transparency and reproducibility. We welcome contributions and
corrections from users. Our package was built using the devtools
methodology (<https://github.com/r-lib/devtools>) in Github. Issues and
feature requests can be contributed in Github
(<https://github.com/trenchproject/TrenchR>). The package is available
via Github and has been submitted to CRAN.

@@

**Utility functions** allow estimating environmental metrics that form
the basis of microclimate and biophysical models (e.g., zenith and
declination angles, which describe the angles of incident sunlight),
facilitate modeling diurnal temperature variation, and estimate
biologically relevant aggregate metrics such as degree days available
for oranismal development.

@@

**Microclimate functions** allow estimating the environmental conditions
experienced by organisms. Temperature and wind profile functions scale
environmental conditions from sensors to organismal height. Radiation
functions allow estimating incoming solar radiation and partitioning the
variation diurnally and across components (i.e., direct, diffuse,
reflected).

@@

We introduce the functions in several vignettes. A good place to start
is the *Allometry and conversions* vignette, which provides tools for
preparing data, such as estimating additional dimensions of organisms
from measured dimensions. The *Estimating microclimates* vignette
provides resources for estimating the environmental conditions
experienced by organisms. This includes estimating solar radiation and
its components, diurnal variation in temperature and radiation,
temperature and wind speed profiles, and soil temperatures and profiles.
Finally, the core biophysical modelling functions are described in a
tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual
functions and then body temperatures can be solved for using either a
generic energy balance or taxa specific biophysical models. We
additionally offer a *List of symbols* used in equations.

For additional background on microclimate and biophysical modelling, we
have updated a series of tutorials entitled *Physical Processes in
Ecosystems*. We intend the tutorials to provide a more contained and
accessible introduction to microclimate and biophysical modelling than
that included in classic textbooks [@campbell2000introduction;
@gates1980biophysical]. The tutorials provide less detailed coverage of
many of the topics contained in the books. The series of 14 tutorials
starts with an overview of the calculus and physics principles
underlying the modelling. Thermodynamics and energy budget modelling are
then introduced with detailed examples. Tutorials address the climate
space concept, operative temperatures, biophysical models for leaves and
sheep, and heat flow in soils. We have revised the tutorials (which
originated from an NSF training grant in 1979 and lacked broad
distribution) to include R code and utilization of TrenchR functions. We
expanded the original series to include tutorials contributed by M.
Kearney introducing the Microclim environmental data
[@kearney2014microclim] and integrating the Microclim data with
biophysical modelling to examine broad scale climatic limits. The
tutorials are provided in html form and available for download as a pdf
at the bookdown server
(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
R markdown files for the tutorials are available in Github
(<https://github.com/trenchproject/TrenchRmodules>).

## Example application

We illustrate the use of the TrenchR package by estimating an energy
budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New
Mexico, USA (35.69°, -105.944°, elevation: 2121 m). The example is
also incorporated in the *Using energy balances to estimate body
temperatures* vignette. We start by generating environmental inputs (Figure 2).
Using these inputs, we estimate the energy budget with component
functions. Finally, we use an integrated biophysical model to estimate
operative environmental temperatures. See Appendix S1 for another
example examining a time series of microclimate data and applying and
testing the biophysical models.

We will use the energy budget to estimate body temperature, $T_b$, which
can depart dramatically from the air temperatures due to energy exchange
with the environment. Heat is gained from absorbed solar and thermal
radiation and from metabolic reactions. Heat is lost through the
organism's emission of radiation and the evaporation of water. The
organism exchanges heat with the surrounding air or water via convection
and with the ground via conduction. The balance of these heat exchanges
(omitting metabolism and evaporation, which are often negligible for
ectotherms) can be estimated and often referred to as operative
environmental temperature, $T_e$ [@bakken1992measurement]. $T_e$ is an
estimate of $T_b$ and the package functions refer to $T_b$ for
simplicity. Our examplar exstimation of $T_b$ assumes thermal equilibrium.

Let us assume the lizard is in an unshaded location where a weather
station at standard height (2 meters) reports that the daily air
temperature varies from a minimum of 10°C to a maximum of 25°C and the
wind speed averages 1 m/s. The soil surface temperature varies from a
minimum of 15°C to a maximum of 30°C. We assume that atmospheric
transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.

@@

```{r}
# Air temperature (˚C)
Ta = sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2) 
# Soil temperature (˚C)
Ts = sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
```

At the second stage, we use microclimate models to scale air temperature and wind speed from
weather station height (2m) to lizard height (0.02m). We assume a
surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and
determines the turbulence of airflow. We implement a neutral air
temperature profile driven by density differences, an air temperature profile forced by wind speed,
and a neutral wind speed profile.

@@

Finally, we will use our microclimates estimates to solve the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
where $Q_{net}$ is the net energy exchange with the environment (W),
$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
the energy emitted due to evaporative water loss (W). We will estimate
each term on the right side of the equation in turn. Estimating
$Q_{abs}$ requires the surface area exposed to radiation and the solar
absorptivity of the animal surface ($a$ proportion). We model a 10 gram
*Sceloporus* lizard with solar absorptivity $a = 0.9$
[@gates1980biophysical]. We will initially assume $T_b = T_a$ to
illustrate the calculations before solving for $T_b$ given the
environmental conditions.

@@

We estimate convective heat exchange between the animal and surrounding
air using the following relationship:
$$Q_{conv} = ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
enhancement factor, $ef$, multiplier can be incorporated to account for
increases in heat exchange resulting from air turbulence in field
conditions. We implement the function in R assuming that 2/3 of the
lizard's surface area is exchanging heat through convection. We
currently estimate no convection since we are assuming $Tb = Ta$ for
illustration purposes:

@@

We estimate conductive heat flow (W) from the lizard to the surface
assuming conductance through the animal tissue is the rate limiting step
as follows:
$$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$ We
implement the estimate in R assuming that 1/3 of the lizard surface is
in contact with the ground and a skin thickness $d = 10^{-6}$:

@@

```{r, echo=FALSE}
par(mar=c(5, 4, 4, 8) + 0.1)
plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="#5DC863FF", ylim=c(10,60)) # Gates
points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") # Campbell and Norman
points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") # Lizard, Buckley 2008
points(1:24, Ta, type="l", col="#440154FF")
points(1:24, Ta_liz, type="l", col="#3B528BFF")
points(1:24, Ts, type="l", col="#21908CFF")

@@

**Figure 3**. Body temperatures (Te) are predicted to drastically exceed
air temperature when lizards are exposed to high levels of solar
radiation. Air temperatures (Ta, °C) at lizard height (0.02m) are
predicted to exceed air temperatures at 2m and to be similar to surface
temperatures (Ts). We estimate body temperatures using two general
energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a
lizard specific biophysical model [dashed: Tb_lizard()] that differ in
how they model heat exchanges.

@@

## Authors' contributions

@@

## Data availability

All code is available at <https://github.com/trenchproject/TrenchR>.

>>>>

1.  Much understanding of organismal responses to climate change and variability relies on the assumption that body temperatures are equal to temporally averaged air temperatures high above the ground. 
    However, most organisms live near the ground and acute exposure to solar and thermal radiation and thermal extremes can substantially elevate their body temperatures.
2.  We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses. 
    The package includes microclimate models to vertically scale weather station data to organismal heights. 
    The biophysical modeling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa. 
    Additional functions model and temporally partition air and soil temperatures and solar radiation. 
    We offer utility functions to aid in estimating the organismal and environmental parameters needed for biophysical ecology.
3.  TrenchR focuses on simple and modular functions so users can create transparent and flexible models for biophysical applications.
4.  The package aims to introduce and enable microclimate and biophysical modeling to improve ecological and evolutionary forecasting. 
    We further this aim through a series of educational modules that introduce the field of biophysical ecology.

@@

**Figure 1** A) Microclimate models scale air and soil temperatures and wind speeds from sensor to organismal heights. B) Biophysical models balance heat exchanges between organisms and their environment to estimate body temperatures [@porter1973behavioral]. A thermal image depicts how grasshopper body temperatures vary substantially from vegetation, air, and ground temperatures due to heat exchange.

@@

## Methods and Features

The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to promote transparency and reproducibility. 
We welcome contributions and corrections from users. 
Our package was built using the devtools methodology (<https://github.com/r-lib/devtools>) with version control managed in Github. 
Issues and feature requests can be contributed in Github (<https://github.com/trenchproject/TrenchR>). 
The package is available via Github and has been submitted to CRAN.

@@

**Utility functions** aid calculation of environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for oranismal development.

@@

**Microclimate functions** facilitate calculating the environmental conditions experienced by organisms. 
Temperature and wind profile functions scale environmental conditions from sensors to organismal height. 
Radiation functions allow estimating incoming solar radiation and partitioning the variation diurnally and across components (i.e., direct, diffuse, reflected).

@@

We introduce the functions in several vignettes. 
A good place to start is the *Allometry and conversions* vignette, which provides tools for preparing data, such as estimating additional dimensions of organisms from measured dimensions. The *Estimating microclimates* vignette provides resources for estimating the environmental conditions experienced by organisms. 
This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and wind speed profiles, and soil temperatures and profiles.
Finally, the core biophysical modeling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models. 
We additionally offer a *List of symbols* used in equations.

For additional background on microclimate and biophysical modeling, we have updated a series of tutorials entitled *Physical Processes in Ecosystems*. 
We intend the tutorials to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical]. 
The tutorials provide less detailed coverage of many of the topics contained in the books. 
The series of 14 tutorials starts with an overview of the calculus and physics principles underlying the modeling. 
Thermodynamics and energy budget modeling are then introduced with detailed examples. 
Tutorials address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils. 
We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions. 
We expanded the original series to include tutorials contributed by M. Kearney introducing the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modeling to examine broad scale climatic limits. 
The tutorials are provided in html form and available for download as a pdf at the bookdown server (<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
R markdown files for the tutorials are available in Github (<https://github.com/trenchproject/TrenchRmodules>).

## Example Application

We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New Mexico, USA (35.69$^\circ$, -105.944$^\circ$, elevation: 2121 m). 
The example is also incorporated in the *Using energy balances to estimate body temperatures* vignette. 
We start by generating environmental inputs (Figure 2).
Using these inputs, we estimate the energy budget with component functions. 
Finally, we use an integrated biophysical model to estimate operative environmental temperatures. 
See Appendix S1 for another example examining a time series of microclimate data and applying and testing the biophysical models.

We will use the energy budget to estimate body temperature, $T_b$, which can depart dramatically from the air temperatures due to energy exchange with the environment. 
Heat is gained from absorbed solar and thermal radiation and from metabolic reactions. 
Heat is lost through the organism's emission of radiation and the evaporation of water. 
The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction. 
The balance of these heat exchanges (omitting metabolism and evaporation, which are often negligible for ectotherms) can be estimated and often referred to as operative environmental temperature, $T_e$ [@bakken1992measurement].
$T_e$ is an estimate of $T_b$ and the package functions refer to $T_b$ for simplicity. 
Our exemplar estimation of $T_b$ assumes thermal equilibrium.

Let us assume the lizard is in an unshaded location where a weather station at standard height (2 meters) reports that the daily air temperature varies from a minimum of 10$^\circ$C to a maximum of 25$^\circ$C and the wind speed averages 1 m/s. 
The soil surface temperature varies from a minimum of 15$^\circ$C to a maximum of 30$^\circ$C. 
We assume that atmospheric transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.

@@

```{r}
# Air temperature ($^\circ$C)
Ta = sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2) 
# Soil temperature ($^\circ$C)
Ts = sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
```

At the second stage, we use microclimate models to scale air temperature and wind speed from weather station height (2 m) to lizard height (0.02 m). 
We assume a surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and determines the turbulence of airflow. 
We implement a neutral air temperature profile driven by density differences, an air temperature profile forced by wind speed, and a neutral wind speed profile.

@@

Finally, we will use our microclimates estimates to solve the following energy balance to estimate $T_e$:

$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$

where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is the energy emitted due to evaporative water loss (W). 
We will estimate each term on the right side of the equation in turn. 
Estimating $Q_{abs}$ requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion). 
We model a 10 gram *Sceloporus* lizard with solar absorptivity $a = 0.9$ [@gates1980biophysical]. 
We will initially assume $T_b = T_a$ to illustrate the calculations before solving for $T_b$ given the environmental conditions.

@@

We estimate convective heat exchange between the animal and surrounding air using the following relationship:

$$Q_{conv} = ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$

where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulence in field conditions. 
We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection. 
We currently estimate no convection since we are assuming $Tb = Ta$ for illustration purposes:

@@

We estimate conductive heat flow (W) from the lizard to the surface assuming conductance through the animal tissue is the rate limiting step as follows:

$$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$

We implement the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thickness $d = 10^{-6}$:

@@

```{r, echo=FALSE}
par(mar=c(5, 4, 4, 8) + 0.1)
plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab= expression("Temperature ("*~degree*C*")"), col="#5DC863FF", ylim=c(10,60)) # Gates
points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") # Campbell and Norman
points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") # Lizard, Buckley 2008
points(1:24, Ta, type="l", col="#440154FF")
points(1:24, Ta_liz, type="l", col="#3B528BFF")
points(1:24, Ts, type="l", col="#21908CFF")

@@

**Figure 3**. Body temperatures (Te) are predicted to drastically exceed air temperature when lizards are exposed to high levels of solar radiation. 
Air temperatures (Ta, $^\circ$C) at lizard height (0.02 m) are predicted to exceed air temperatures at 2 m and to be similar to surface temperatures (Ts). 
We estimate body temperatures using two general energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges.

@@

## Authors' Contributions

@@

## Data Availability

All code and data are available at <https://github.com/trenchproject/TrenchR>.
